A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-26 text/abstract.Rmd
incorporate Megan Hitchens' comments, fix spaces after parens

Cancer is a complex disease involving dysfunctional cellular metabolism, other
basic biochemical processes, and the microenvironment. However, despite some
generally agreed upon unifying principles [@weinberg2000; @hallmarks_II],
molecular signatures remain largely indistinguishable from tissue-of-origin,
presenting a major barrier for precision health and individualized medicine. To
address this challenge, we reduced mutation data to disruptions in a select set
of pathways relevant to basic cell biology, from DNA replication to cellular
communication. Using dimensionality reduction techniques, we assigned tumor
samples into ten clusters distinct from their tissues-of-origin and largely free
of bias from mutational burden or clinical stage. We show that these tumor
classes vary in prognosis by modeling relative risk of death in different
classes and cancer types. We identify cluster-specific mutations in different
tissues, demonstrating that tissue-specific signatures contribute to common
cellular phenotypes. Moreover, driver gene mutations involved in replication
fidelity and genome stability, such as _BRCA1_ and _Rb1_, are equally
distributed among our clusters, contrary to the expectation that such genes are
avatars of molecular subtype. We also investigate pathways that distinguish
between metastatic and non-metastatic tumors, and show that most meaningful
differences are cluster-specific. Some metastasis enriched pathways from one
cluster are cluster-specific pathways from non-metastatic tumors of another
cluster, suggesting phenotypic convergence, and potentially highlighting key
pathways to focus on for prevention of cluster-specific relapse. Taken as a
whole, our observations suggest that common driver genes combine with
tissue-specific disruptions in tumor-promoting pathways to produce a limited
number of distinct molecular phenotypes. Eventually these signatures converge on
one or more end-stage disease phenotypes. Thus, we present a coherent view of
global tumor biology, and explain how common cellular dysfunction might arise
from tissue-specific mutations.

>>>>

Cancer is a complex disease involving disrupted cellular metabolism, basic
biochemical processes, and the microenvironment. However, despite some generally
agreed upon unifying principles [@weinberg2000; @hallmarks_II], molecular
signatures remain largely indistinguishable from tissue-of-origin, presenting a
major barrier for precision health and individualized medicine. To address this
challenge, we reduce mutation data to disruptions in a select set of pathways
relevant to basic cell biology, from DNA replication to cellular communication.
Using dimensionality reduction techniques, we assign tumor samples into ten
clusters distinct from tissue-of-origin and largely free of bias from mutational
burden or clinical stage. We show that the clusters vary in prognosis by
modeling relative risk of death by cancer type and cluster. We identify
cluster-specific mutations in different tissues, demonstrating that
tissue-specific signatures contribute to common cellular phenotypes. Moreover,
germline risk genes involved in replication fidelity and genome stability are
equally distributed among clusters, contrary to the expectation that such genes
are avatars of molecular subtype. We investigate metastatic and non-metastatic
pathways, and show that most differences are cluster-specific. Some metastatic
pathways from one cluster are cluster-specific pathways from non-metastatic
tumors of another cluster, suggesting phenotypic convergence. Taken as a whole,
our observations suggest that common driver genes combine with tissue-specific
disruptions in tumor-promoting pathways to produce a limited number of distinct
molecular phenotypes. Thus, we present a coherent view of global tumor biology,
and explain how common cellular dysfunction might arise from tissue-specific
mutations.

### 2020-08-26 text/discussion.Rmd
incorporate Megan Hitchens' comments, fix spaces after parens

One of the biggest hurdles in cancer research is the sparsity of data; ~20,000
protein-coding genes is comparable with the number of tumor samples, even with
multiple mutations per sample. And that is not accounting for the fact that
~30,000 non-coding genes and hundreds of thousands of non-coding regulatory
elements exist with oncogenic or tumor suppressor activity [@deesMuSiC2012;
@tamboreroOncodriveCLUST2013; @lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
@zhaoModelingPositive2019; @noncoding-mutations-Zhang]. Hence, we sought to
simplify the problem by employing a "knowledge-base driven analysis"
[@pathways-Atul-Butte], investigating cancer as a disease of basic cellular and
biochemical pathways, which we accomplished by translating gene-level mutations
into pathway level disruptions. Our approach differs from previously described
methods [@PATHNET] in that we chose to focus on the pathway (defined in the
methods) as the unit of disruption instead of the gene, employing the conceit
that individual mutations may be sufficient to disrupt or misdirect pathway
activity (as is often the case with driver mutations). Rather than evaluating
the burden of mutations on each pathway within or across tumors, which can be
biased by the number of genes or the coverage of pathway-linked genes in the
genome, we scored each pathway in binary fashion, allowing one mutation or
multiple mutations to count equally for each sample.

To our knowledge, this approach has not previously been attempted in spite of
its relative simplicity. We limited our analysis to mutations with likely
deleterious effects, including missense, nonsense, frame-shift, stop-loss,
untranslated region and splice site mutations in genes that are actively
expressed in each cancer type, thus avoiding bias from transcription coupled
repair. Our method of filtration differs from the "rank-and-cut" method of
[@PCAWG-MAIN] but represents a reasonable attempt to account for the same
biases. We also filtered genes on the basis of pathway membership in biochemical
pathways, excluding curated gene sets related to diseases, syndromes, or classes
of proteins with shared catalytic activity or conserved domains which are
potentially problematic [@pathways-Atul-Butte]. We chose this approach to limit
redundancy and exclude biologically unrelated collections of genes. Our
pathway-centric analysis allows us to circumvent the problem that some
cancer-promoting mutations likely occur at such low frequencies as to be
indistinguishable from background. Previous attempts to capture these
associations using Bayesian networks have yielded important new insights into
this problem and new genes [@hornNetSigNetworkbasedDiscovery2018] that would not
have been possible with traditional analysis, supporting the case that
pathway-centric approaches have the potential to increase statistical power.

@@

We find that most tumors segregate into one of ten biologically distinct
clusters, regardless of the cancer type. Some cancer types are unevenly
distributed, though we could not identify any cancers that were exclusive to a
single tumor cluster. Only cancer types with the fewest samples were found to be
absent from one or more clusters at all. Somewhat surprisingly to us, this
finding extends to histological subtypes of breast, head and neck cancers,
leukemias, _etc._ This result implies that many, if not most, histological
subtypes could reflect differences in cell-of-origin, rather than fundamental
differences in proximal cause. The four major subtypes of breast cancer
correspond to histological and molecular expression profiles that define them
and how they respond to experimental stimuli [@breast-subtypes2]. It has been
hypothesized that differences in the molecular regulators of development in the
precursor cell types present in the breast epithelium drive the various
histological phenotypes [@breast-subtypes; @breast-subtypes3]. Consistent with
this view we found that breast tumor samples of the Luminal A subtype were
heavily biased toward membership in clusters 1 and 7, and basal tumor samples
were biased toward cluster 6. However, both subtypes also contained samples in
every other cluster (without exception), and Luminal B and Her2 positive samples
are distributed across clusters. We interpret these data to mean that inherited
cell-of-origin molecular profiles may predispose certain precursor cell types
within the breast epithelium to forming tumors of one cluster or another but are
not determinative; this view is compatible with the previously stated hypothesis
but opens the way for a more granular view of individual tumors. It would be
surprising if we did not observe bias for some cancers and subtypes amongst our
classes, since some treatment regimens have greater efficacy for patients of a
given cancer or histological subtype [@breast-subtypes2]. Nonetheless, the basis
for some tumors being treatment-refractory in spite of receiving the standard of
clinical care for diagnostic markers remains elusive. Doubtless some of this is
due to bad luck, as tumors can metastasize and remain dormant years before they
are detected at distal sites, or resistant clones may have already arisen at
undetectable levels [@hallmarks_II], but our analysis suggests the possibility
of identifying more informative molecular, histological or cellular subtypes
that could form a basis for future separation of patients into different
precision treatment regimens.

>>>>

One of the biggest hurdles in cancer research is the sparsity of data; ~20,000
protein-coding genes is comparable with the number of tumor samples, even with
multiple mutations per sample. This also does not account for the fact that
~30,000 non-coding genes and hundreds of thousands of non-coding regulatory
elements exist with oncogenic or tumor suppressor activity [@deesMuSiC2012;
@tamboreroOncodriveCLUST2013; @lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
@zhaoModelingPositive2019; @noncoding-mutations-Zhang]. Hence, we sought to
simplify the problem by employing a "knowledge-base driven analysis"
[@pathways-Atul-Butte], investigating cancer as a disease of basic cellular and
biochemical pathways, which we accomplished by translating gene-level mutations
into pathway level disruptions. Our approach differs from previously described
methods [@PATHNET] in that we chose to focus on the pathway (defined in the
methods) as the unit of disruption instead of the gene, where individual
mutations may be sufficient to disrupt or misdirect pathway activity (as is
often the case with driver mutations). Rather than evaluating the burden of
mutations on each pathway within or across tumors, which can be biased by the
number of genes or the coverage of pathway-linked genes in the genome, we scored
each pathway in binary fashion, allowing one mutation or multiple mutations to
count equally for each sample.

To our knowledge, this approach has not previously been attempted in spite of
its relative simplicity. We limited our analysis to mutations with likely
deleterious effects, including missense, nonsense, frame-shift, stop-loss,
untranslated region and splice site mutations in genes that are actively
expressed in each cancer type, thus avoiding bias from transcription coupled
repair. Our method of filtration differs from the "rank-and-cut" method of
[@PCAWG-MAIN] but represents a reasonable attempt to account for the same
biases. We also filtered genes on the basis of pathway membership in biochemical
pathways, excluding curated gene sets related to diseases, syndromes, or classes
of proteins with shared catalytic activity or conserved domains which are
potentially problematic [@pathways-Atul-Butte]. We chose this approach to limit
redundancy and exclude biologically unrelated collections of genes. Our
pathway-centric analysis allows us to circumvent the problem that some
cancer-promoting mutations likely occur at such low frequencies as to be
indistinguishable from background.

@@

We find that most tumors segregate into one of ten biologically distinct
clusters, regardless of the cancer type. Some cancer types are unevenly
distributed, though we could not identify any cancers that were exclusive to a
single tumor cluster. Only cancer types with the fewest samples were found to be
absent from one or more clusters at all. Somewhat surprisingly to us, this
finding extends to histological subtypes of breast, head and neck cancers,
leukemias, _etc._ This result implies that many, if not most, histological
subtypes could reflect differences in cell-of-origin, rather than fundamental
differences in proximal cause. The four major subtypes of breast cancer
correspond to histological and molecular expression profiles that define them
and how they respond to experimental stimuli [@breast-subtypes2]. It has been
hypothesized that differences in the molecular regulators of development in the
precursor cell types present in the breast epithelium drive the various
histological phenotypes [@breast-subtypes; @breast-subtypes3]. Consistent with
this view we found that breast tumor samples of the Luminal A subtype were
heavily biased toward membership in clusters 1 and 7, and basal tumor samples
were biased toward cluster 6. However, both subtypes also contained samples in
every other cluster (without exception), and Luminal B and Her2 positive samples
are distributed across clusters. We interpret these data to mean that inherited
cell-of-origin molecular profiles may predispose certain precursor cell types
within the breast epithelium to forming tumors of one cluster or another but are
not determinative; this view is compatible with the previously stated hypothesis
but opens the way for a more granular view of individual tumors. It would be
surprising if we did not observe bias for some cancers and subtypes amongst our
classes, since some treatment regimens have greater efficacy for patients of a
given cancer or histological subtype [@breast-subtypes2]. Nonetheless, the basis
for some tumors being treatment-refractory in spite of receiving the standard of
clinical care for diagnostic markers remains elusive. Doubtless some of this is
due to chance events, as tumors can metastasize and remain dormant years before
they are detected at distal sites, or resistant clones may have already arisen
at undetectable levels [@hallmarks_II], but our analysis suggests the
possibility of identifying more informative molecular, histological or cellular
subtypes that could form a basis for future stratification of patients into
different precision treatment regimens.

### 2020-08-26 text/introduction.Rmd
incorporate Megan Hitchens' comments, fix spaces after parens

In this study, we attempt to address these questions by focusing on discrete
molecular pathways as the unit of disruption in cancer. Our approach strips all
cell-type-specific information from the mutation data and equates gene-level
mutations to cell-biological pathway disruptions. We use this heuristic to
evaluate all cancers and show that, surprisingly, tumors that exhibit
tissue-specific gene mutation patterns nonetheless fall into common categories
of pathway disruption having unique prognoses in each cancer type.

>>>>

In this study, we attempt to understand how tissue-specific gene disruptions
create common cancer phenotypes by focusing on discrete molecular pathways as
the unit of disruption. Our approach strips all cell-type-specific information
from the mutation data and equates gene-level mutations to cell-biological
pathway disruptions. We use this heuristic to evaluate all cancers and show
that, surprisingly, tumors that exhibit tissue-specific gene mutation patterns
nonetheless fall into common categories of pathway disruption having unique
prognoses in each cancer type.

### 2020-08-26 text/manuscript.Rmd
incorporate Megan Hitchens' comments, fix spaces after parens

---
title: "A molecular taxonomy of tumors independent of tissue-of-origin"
author: Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and
  Dennis J. Hazelett$^{1,3}$
bibliography: references.bib
output:
  word_document:
    reference_docx: "../text/word-style-reference-ttmanu.docx"
always_allow_html: TRUE
editor_options:
  chunk_output_type: console
---

###### Institutional affiliations: 1. The Center for Bioinformatics and Functional Genomics, Cedars-Sinai Medical Center, Los Angeles, California. 2. Lakeland Applied Sciences LLC, Los Angeles, CA. 3. Samuel Oschin Comprehensive Cancer Institute, Cedars-Sinai Medical Center, Los Angeles, CA.

@@

To study cancer pathways we obtained a set of 7,607 solid tumor samples from The
Cancer Genome Atlas (TCGA) through the [Genomic Data Commons
(GDC)](gdc.cancer.gov) portal. TCGA data were most appropriate for our study
given the relative completeness of the patient metadata, particularly for
survival and staging. We chose to analyze somatic mutations in exome sequencing
data because the affected target gene is known unambiguously. Therefore, we
selected all missense, nonsense, frameshift, stop-loss, untranslated region, and
splicing mutations. In order to minimize bias from well-studied diseases and
processes, we selected 377 [Reactome pathways](https://reactome.org/) (see
supplemental table ST1) of interest corresponding to basic cellular processes
and biochemical pathways, excluding gene sets that correspond to miscellaneous
categories ( _e.g._ “transcription factors”) or disease associations ( _e.g._
“mutated in colon cancer”) and filtered our gene list on membership in these
pathways (total of 8,940 genes; see methods for details).

To avoid bias toward larger pathways ( _i.e._ pathways with more member genes),
we counted pathways as disrupted if one or more member genes were mutated. We do
not attempt to calculate enrichment for mutations within a pathway. Data
binarized by pathway are likely to be noisy for several reasons. First, point
mutations can be deleterious (attenuating or hypomorphic) or activating
(neomorphic or hypermorphic) in genes, and these can in turn act functionally as
either activators or repressors. For this study we assume a significant fraction
of these mutations are generically disruptive to normal pathway activity if they
are activating or deactivating or affect a gene that one might classify as a
tumor suppressor or oncogene. Second, we know that low-expressed genes and
non-expressed genes accumulate mutations at a higher rate due to transcription
coupled repair [@kandothMutationalLandscapeSignificance2013;
@kimPancancerAnalysisSomatic2018;
@Pervasive_lesion_segregation; @expression-aware-annotation]. To address this
issue, we identified genes with low expression in each type of cancer and
eliminated them for that cancer type only (see methods for details). Highly
expressed genes could also have high mutation rates owing to transcription
induced mutagenesis [@parkGenomicEvidenceElevated2012]. We felt that such
mechanisms result in cell-type-specific biases that might be biologically
meaningful for predisposition to different classes of cancer however, and
therefore chose not to exclude these genes from our analysis. After selecting
our pathways and genes, we then compiled a matrix of the pathways, assigning a
Boolean value of 1 to each pathway with one or more genes mutated and 0 for all
others (**figure 1**).

@@

Tumor staging is based on physico-pathological criteria, including tumor
diameter, which can vary greatly in importance between different tissues. Stage
is used clinically as a proxy for advancement toward a more deadly state and
metastasis. Given these assumptions, it is possible that more advanced tumors
have common pathway disruption profiles, and the UMAP which features a series of
lobe-like structures on a common backbone of tumor samples (**figure 2**) could
in principle also reflect progression through a series of stages. In support of
this hypothesis, the backbone starts with a cluster of tumors (class 1) that has
the fewest point mutations and culminates in a cluster (class 10) which has
nearly every pathway disrupted (**figure 4**). However, outside of class 10 we
don't observe an obvious trend in the overall mutation burden across the
backbone of the UMAP. Nonetheless, to test the hypothesis that
the molecular-pathway disruption clusters represent advancement through stages,
we projected staging data onto the UMAP. Similar to tissue of origin and other
categories of tumor, we did not observe any bias among the stages to specific
pathway disruption clusters (**figure 3D**), suggesting that stage is not a
contributing factor to cluster identity.

@@

To discover what pathways are most important for clustering, we calculated
percent enrichment for each pathway _within cluster_ relative to all other
clusters combined (see methods) and ranked pathways for each cluster from
highest to lowest enrichment. We visualized the enrichment as a heatmap
(**figure 5A**). Using this approach, we identified 25 pathways highly enriched
(enrichment score ≥ 0.3, 95% confidence; see methods) in cluster 2, 9 pathways
enriched in cluster 3, 24 pathways in cluster 4, 10 pathways in cluster 5, 10
pathways in cluster 6, 24 pathways in cluster 7, 24 pathways in cluster 8, 10
pathways in cluster 9, and 0 pathways for clusters 1 and 10 (**supplementary table ST2**). Clusters 7, 8, and
9 had several pathways in common. To explore the specific pathways marking each
cluster, we projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4**). Clusters 3 and 5 were distinguished by metabolic
pathways including RNA and protein biosynthesis (**supplementary figure S4**).
Similarly, cluster 4 was distinguished by mutations affecting regulation of DNA
and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
and "Nucleosome assembly"). Clusters 7-9 have in common mutations in
extracellular, intracellular, and immune-related signaling pathways (see
**figure 5B** and **supplementary figure S4**). Cluster 2 had the highest
pathway enrichment levels of the three, having mutations in hedgehog signaling,
"β-catenin degradation", "cellular response to hypoxia", "regulation of cell
cycle" and "apoptosis" among others.

@@

Prior efforts to extract common signatures from pan-cancer datasets met with
difficulty in distinguishing tumor samples from their tissue-specific -omics
data signatures. Given our pathway-disruption based clustering, this raises
the question, are tumor phenotypes driven entirely by common driver genes, or by
“silent” tissue-specific effectors ( _i.e._ too few samples to detect above
statistical significance thresholds), or a combination of both? To answer this
question, we compared top pathway genes for each cluster relative to the TCGA
background to find differentially mutated genes. We ranked odds ratios and
selected the top ten enriched and depleted genes (pvalue < 0.01) for each
cluster (**Figure 6**; odds ratios plot). Clusters 7 and 8, which shared
multiple enrichment in signaling pathways, are largely driven by mutations in
_PI3K_ and its orthologs and _Ras_ genes, respectively (compare _PIK3CA_ and
_KRAS_ panels of **Supplementary figure S5**). Interestingly, cluster 9, which
also shared multiple enrichment in signaling pathways with clusters 7 and 8, is
enriched for both _PIK3CA_ and _KRAS_. Clusters 3 and 5, defined by enrichment
in metabolic pathways, had mutations in ribosomal proteins and nuclear pore
complex, respectively. Cluster 4 had mutations in genes responsible for
nucleosome structure. Cluster 2 had mutations in proteasomal subunit genes
involved in protein degradation. We also observed that genes that were enriched
for one cluster are depleted from others ( _e.g. TP53_ is enriched in cluster 6,
but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in
clusters 3 and 8). Next, we investigated the proportion of samples per cancer
type for the significant genes within a cluster (**Figure 6**; heatmap).
Surprisingly, cluster-specific tumors were not predominated by one or more
highly mutated genes across all cancers. Instead, when observing the mutation
rate for these genes within samples that belong to a cluster, the mutation rate
is heterogeneous across tumors by tissue origin ( _e.g._ In cluster 4, CESC was
enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was enriched for
_HIST1H2AC_). Even among the top most mutationally enriched genes within
clusters there is no global pattern, indicating that our clusters are not driven
by individual genes, but rather networks as a whole. Taken together, our data
identifies a framework of cancer type-specific mutations associated with
specific clusters.

@@

Following the same logic we used to investigate cluster-specific enrichment of
pathways we compared metastatic _vs._ non-metastatic tumors, as we find that
metastatic tumors are distributed across all ten clusters ( _e.g._ **figure
3E**). Using all non-metastatic tumors as background, we found very low levels
of enrichment (< 10%) in a handful of pathways. We reasoned that the individual
clusters might be too different to detect global metastasis enrichment signal
given the small sample size (n = 215 metastatic tumor samples).

@@

To understand the molecular mechanism of cancer at a pathway level, we used
Reactome (https://reactome.org/), a knowledge-based pathway database. The
mapping files of ENSEMBL genes to pathways, pathway hierarchy relationships, and
complexes to top pathways were downloaded from
https://reactome.org/download-data. Using these data, we imposed pathway
criteria to define basic cellular processes and biochemical pathways: (1)
human-derived pathways ( _“HSA”_) (2) limited to grandchild node for each parent
pathway ( _e.g._ ‘Beta-catenin independent WNT signaling’ in ‘Signal
Transduction’) (3) exclusion of pathways in the parent pathway: "Disease",
"Muscle contraction", and "Reproduction" or pathway names that include any of
the following keywords: "disease", "defect", "cancer", "oncogenic", "disorder",
"mutant", "loss", "infection", "bacteria", or "listeria". While some of the
excluded pathways have been shown to play an important role in cancer, they are
highly specialized ( _e.g._ "PI3K/AKT Signaling in Cancer"). Additionally, for
most of the excluded pathways, a neutral version pathway of the pathway exists (
_e.g._ "PIP3 activates Akt signaling"). Finally, we mapped the 18,577 Ensembl
IDs from the TCGA dataset to the highly selected Reactome pathways. This
operation produced a lookup table that consisted of 377 pathways mapped to 8,940
genes.

@@

We filtered likely erroneous mutations due to transcription coupled repair. Our
approach was to determine the status of each gene ( _i.e._ expressed or not
expressed) in each tissue type in order to exclude gene mutations low expressed
enes. To do this, we obtained the TCGA RNA sequencing data adjusted for batch
effect dataset (https://pancanatlas.xenahubs.net). Using the data, we removed
the genes and tumor samples that were not included in our analysis, grouped the
tumor samples by tissue type and computed the mean expression value for each
gene. A minimum threshold of 10 transcripts per million was set for expressed
genes based on an inflection point observed when plotting the mean expression
values of genes ranked by expression in each tissue type. Genes that did not
meet this threshold were considered not expressed. This operation produced a
lookup table for gene expression status in each tumor sample for 18,127 genes.

@@

We wish to thank David Van Valen, Simon Knott and Kate Lawrenson for critical
reading of this manuscript, and Ivetth Corona for early discussions and
feedback.

>>>>

---
title: "A molecular taxonomy of tumors independent of tissue-of-origin"
author: Peter T. Nguyen$^{1,\dag}$, Simon G. Coetzee$^{1,\dag}$, Daniel L. Lakeland$^{2}$, and
  Dennis J. Hazelett$^{1,3,*}$
bibliography: references.bib
output:
  word_document:
    reference_docx: "../text/word-style-reference-ttmanu.docx"
always_allow_html: TRUE
editor_options:
  chunk_output_type: console
---

###### Institutional affiliations: 1. The Center for Bioinformatics and Functional Genomics, Cedars-Sinai Medical Center, Los Angeles, California. 2. Lakeland Applied Sciences LLC, Los Angeles, CA. 3. Samuel Oschin Comprehensive Cancer Institute, Cedars-Sinai Medical Center, Los Angeles, CA. $\dag$ These authors contributed equally. $*$ To whom correspondence should be addressed.

@@

To study cancer pathways we obtained a set of 7,607 solid tumor samples from The
Cancer Genome Atlas (TCGA) through the [Genomic Data Commons
(GDC)](gdc.cancer.gov) portal. TCGA data were most appropriate for our study
given the relative completeness of the patient metadata, particularly for
survival and staging. We chose to analyze somatic mutations in exome sequencing
data because the affected target gene is known unambiguously. Therefore, we
selected all missense, nonsense, frameshift, stop-loss, untranslated region, and
splicing mutations. In order to minimize bias from well-studied diseases and
processes, we selected 377 [Reactome pathways](https://reactome.org/) (see
supplemental table ST1) of interest corresponding to basic cellular processes
and biochemical pathways, excluding gene sets that correspond to miscellaneous
categories (\it{e.g.} “transcription factors”) or disease associations (\it{e.g.}
“mutated in colon cancer”) and filtered our gene list on membership in these
pathways (total of 8,940 genes; see methods for details).

To avoid bias toward larger pathways (\it{i.e.} pathways with more member genes),
we counted pathways as disrupted if one or more member genes were mutated. We do
not attempt to calculate enrichment for mutations within a pathway. Data
binarized by pathway are likely to be noisy for several reasons. First, point
mutations can be deleterious (attenuating, hypomorphic or antimorphic) or
activating (neomorphic or hypermorphic) in genes, and these can in turn act
functionally as either oncogenes or tumor suppressors. For this study we assume
a significant fraction of these mutations are generically disruptive to normal
pathway activity since it is impossible to know the tumor promoting effects of
all mutations, including rarely studied genes. Second, we know that
low-expressed genes and non-expressed genes accumulate mutations at a higher
rate due to transcription coupled repair
[@kandothMutationalLandscapeSignificance2013;
@kimPancancerAnalysisSomatic2018;
@Pervasive_lesion_segregation; @expression-aware-annotation]. To address this
issue, we identified genes with low expression in each type of cancer and
eliminated them for that cancer type only (see methods for details). Highly
expressed genes could also have high mutation rates owing to transcription
induced mutagenesis [@parkGenomicEvidenceElevated2012]. We felt that such
mechanisms result in cell-type-specific biases that might be biologically
meaningful for predisposition to different classes of cancer however, and
therefore chose not to exclude these genes from our analysis. After selecting
our pathways and genes, we then compiled a matrix of the pathways, assigning a
Boolean value of 1 to each pathway with one or more genes mutated and 0 for all
others (**figure 1**).

@@

Tumor staging is based on physico-pathological criteria, including tumor
diameter, which can vary greatly in importance between different tissues. Stage
is used clinically as a proxy for advancement toward a more deadly state and
metastasis. Given these assumptions, it is possible that more advanced tumors
have common pathway disruption profiles. The UMAP, which features a series of
lobe-like structures on a common backbone of tumor samples (**figure 2**) could
in principle reflect progression through a series of stages. The backbone starts
with a cluster of tumors (class 1) that has the fewest point mutations and
culminates in a cluster (class 10) which has nearly every pathway disrupted
(**figure 4**). However, outside of class 10 we don't observe an obvious trend
in the overall mutation burden across the backbone of the UMAP. Nonetheless, to
test the hypothesis that the molecular-pathway disruption clusters represent
advancement through stages, we projected staging data onto the UMAP. Similar to
tissue of origin and other categories of tumor, we did not observe any bias
among the stages to specific pathway disruption clusters (**figure 3D**),
suggesting that stage is not a contributing factor to cluster identity.

@@

To discover what pathways are most important for clustering, we calculated
percent enrichment for each pathway _within cluster_ relative to all other
clusters combined (see methods) and ranked pathways for each cluster from
highest to lowest enrichment. We visualized the enrichment as a heatmap
(**figure 5A**). Using this approach, we identified twenty-five pathways highly
enriched (enrichment score ≥ 0.3, 95% confidence; see methods) in cluster 2,
nine pathways enriched in cluster 3, twenty-four pathways in cluster 4, ten
pathways in cluster 5, ten pathways in cluster 6, twenty-four pathways in
cluster 7, twenty-four pathways in cluster 8, ten pathways in cluster 9, and no
pathways for clusters 1 and 10 (**supplementary table ST2**). Clusters 7, 8, and
9 had several pathways in common. To explore the specific pathways marking each
cluster, we projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4**). Clusters 3 and 5 were distinguished by metabolic
pathways including RNA and protein biosynthesis (**supplementary figure S4**).
Similarly, cluster 4 was distinguished by mutations affecting regulation of DNA
and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
and "Nucleosome assembly"). Clusters 7-9 have in common mutations in
extracellular, intracellular, and immune-related signaling pathways (see
**figure 5B** and **supplementary figure S4**). Cluster 2 had the highest
pathway enrichment levels of the three, having mutations in hedgehog signaling,
"β-catenin degradation", "cellular response to hypoxia", "regulation of cell
cycle" and "apoptosis" among others.

@@

Prior efforts to extract common signatures from pan-cancer datasets met with
difficulty in distinguishing tumor samples from their tissue-specific -omics
data signatures. Given our pathway-disruption based clustering, this raises
the question, are tumor phenotypes driven entirely by common driver genes, or by
“silent” tissue-specific effectors (\it{i.e.} too few samples to detect above
statistical significance thresholds), or a combination of both? To answer this
question, we compared top pathway genes for each cluster relative to the TCGA
background to find differentially mutated genes. We ranked odds ratios and
selected the top ten enriched and depleted genes (pvalue < 0.01) for each
cluster (**Figure 6**; odds ratios plot). Clusters 7 and 8, which shared
multiple enrichment in signaling pathways, are largely driven by mutations in
_PI3K_ and its orthologs and _Ras_ genes, respectively (compare _PIK3CA_ and
_KRAS_ panels of **Supplementary figure S5**). Interestingly, cluster 9, which
also shared multiple enrichment in signaling pathways with clusters 7 and 8, is
enriched for both _PIK3CA_ and _KRAS_. Clusters 3 and 5, defined by enrichment
in metabolic pathways, had mutations in ribosomal proteins and nuclear pore
complex, respectively. Cluster 4 had mutations in genes responsible for
nucleosome structure. Cluster 2 had mutations in proteasomal subunit genes
involved in protein degradation. We also observed that genes that were enriched
for one cluster are depleted from others (\it{i.e.} is enriched in cluster 6,
but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in
clusters 3 and 8). Next, we investigated the proportion of samples per cancer
type for the significant genes within a cluster (**Figure 6**; heatmap).
Surprisingly, cluster-specific tumors were not predominated by one or more
highly mutated genes across all cancers. Instead, when observing the mutation
rate for these genes within samples that belong to a cluster, the mutation rate
is heterogeneous across tumors by tissue origin (\it{e.g.} In cluster 4, CESC was
enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was enriched for
_HIST1H2AC_). Even among the top most mutationally enriched genes within
clusters there is no global pattern, indicating that our clusters are not driven
by individual genes, but rather networks as a whole. Taken together, our data
identifies a framework of cancer type-specific mutations associated with
specific clusters.

@@

Following the same logic we used to investigate cluster-specific enrichment of
pathways we compared metastatic _vs._ non-metastatic tumors, as we find that
metastatic tumors are distributed across all ten clusters (\it{e.g.} **figure
3E**). Using all non-metastatic tumors as background, we found very low levels
of enrichment (< 10%) in a handful of pathways. We reasoned that the individual
clusters might be too different to detect global metastasis enrichment signal
given the small sample size (n = 215 metastatic tumor samples).

@@

To understand the molecular mechanism of cancer at a pathway level, we used
Reactome (https://reactome.org/), a knowledge-based pathway database. The
mapping files of ENSEMBL genes to pathways, pathway hierarchy relationships, and
complexes to top pathways were downloaded from
https://reactome.org/download-data. Using these data, we imposed pathway
criteria to define basic cellular processes and biochemical pathways: (1)
human-derived pathways (\it{“HSA”}) (2) limited to grandchild node for each
parent pathway (\it{e.g.} ‘Beta-catenin independent WNT signaling’ in ‘Signal
Transduction’) (3) exclusion of pathways in the parent pathway: "Disease",
"Muscle contraction", and "Reproduction" or pathway names that include any of
the following keywords: "disease", "defect", "cancer", "oncogenic", "disorder",
"mutant", "loss", "infection", "bacteria", or "listeria". While some of the
excluded pathways have been shown to play an important role in cancer, they are
highly specialized (\it{e.g.} "PI3K/AKT Signaling in Cancer"). Additionally, for
most of the excluded pathways, a neutral version pathway of the pathway exists
(\it{e.g.} "PIP3 activates Akt signaling"). Finally, we mapped the 18,577
Ensembl IDs from the TCGA dataset to the highly selected Reactome pathways. This
operation produced a lookup table that consisted of 377 pathways mapped to 8,940
genes.

@@

We filtered likely erroneous mutations due to transcription coupled repair. Our
approach was to determine the status of each gene (\it{i.e.} expressed or not
expressed) in each tissue type in order to exclude gene mutations low expressed
enes. To do this, we obtained the TCGA RNA sequencing data adjusted for batch
effect dataset (https://pancanatlas.xenahubs.net). Using the data, we removed
the genes and tumor samples that were not included in our analysis, grouped the
tumor samples by tissue type and computed the mean expression value for each
gene. A minimum threshold of 10 transcripts per million was set for expressed
genes based on an inflection point observed when plotting the mean expression
values of genes ranked by expression in each tissue type. Genes that did not
meet this threshold were considered not expressed. This operation produced a
lookup table for gene expression status in each tumor sample for 18,127 genes.

@@

We wish to thank David Van Valen, Kate Lawrenson, Simon Knott, and Megan
Hitchens for critical reading of this manuscript, and Ivetth Corona for early
discussions and feedback.
